---
output: html_document
editor_options: 
  chunk_output_type: console
---
White Wine Exploration by Kevin Foong
========================================================
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

library(ggplot2)
library(gridExtra)
library(dplyr)
library(GGally)
library(RColorBrewer)
library(memisc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wines = read.csv('wineQualityWhites.csv')
```

This dataset contains 4,898 white wines with 11 variables on various chemical 
properties of each wine. At least 3 wine experts rated the quality of each wine,
providing a rating between 0 (very bad) and 10 (excellent).

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
print (c(nrow(wines),ncol(wines)))
str(wines)
summary(wines)
```

This dataset consists of 13 variables including one index which we will ignore 
leaving us with 11 chemical properties and 1 quality rating.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_histogram}
p1 = ggplot(data=wines,aes(x=fixed.acidity))+
  geom_histogram(binwidth=0.2)

p2 = ggplot(data=wines,aes(x=volatile.acidity))+
  geom_histogram(binwidth=.02)

p3 = ggplot(data=wines,aes(x=citric.acid))+
  geom_histogram(binwidth=.02)

p4 = ggplot(data=wines,aes(x=residual.sugar))+
  geom_histogram(binwidth=1)

p5 = ggplot(data=wines,aes(x=chlorides))+
  geom_histogram(binwidth=.004)

p6 = ggplot(data=wines,aes(x=free.sulfur.dioxide))+
  geom_histogram(binwidth=2)

p7 = ggplot(data=wines,aes(x=total.sulfur.dioxide))+
  geom_histogram(binwidth=5)

p8 = ggplot(data=wines,aes(x=density))+
  geom_histogram(binwidth=.0005)

p9 = ggplot(data=wines,aes(x=pH))+
  geom_histogram(binwidth=.02)

p10 = ggplot(data=wines,aes(x=sulphates))+
  geom_histogram(binwidth=.02)

p11 = ggplot(data=wines,aes(x=alcohol))+
  geom_histogram(binwidth=.2)

p12 = ggplot(data=wines,aes(x=quality))+
  geom_bar()

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol=3)
```

To get an overview of the data we first take a look at the histograms of all 12
variables, experimenting with the binwidth to capture the right granularity. 
Alot of variables are long-tailed with high values. To remove the outliers I 
will replot the histograms, omitting the top 1% of data of those variables which
are long-tailed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_histogram_2}
p1 = ggplot(data=wines,aes(x=fixed.acidity))+
  geom_histogram(binwidth=0.2)+
  xlim(c(3.8,quantile(wines$fixed.acidity,0.99)))

p2 = ggplot(data=wines,aes(x=volatile.acidity))+
  geom_histogram(binwidth=.02)+
  xlim(c(0.08,quantile(wines$volatile.acidity,0.99)))  

p3 = ggplot(data=wines,aes(x=citric.acid))+
  geom_histogram(binwidth=.02)+
  xlim(0,quantile(wines$citric.acid,0.99))

p4 = ggplot(data=wines,aes(x=residual.sugar))+
  geom_histogram(binwidth=1)+
  xlim(c(0.6,quantile(wines$residual.sugar,0.99))) 

p5 = ggplot(data=wines,aes(x=chlorides))+
  geom_histogram(binwidth=.004)+
  xlim(c(0,quantile(wines$chlorides,0.99))) 

p6 = ggplot(data=wines,aes(x=free.sulfur.dioxide))+
  geom_histogram(binwidth=2)+
  xlim(c(0.009,quantile(wines$free.sulfur.dioxide,0.99)))

p7 = ggplot(data=wines,aes(x=total.sulfur.dioxide))+
  geom_histogram(binwidth=5)+
  xlim(c(0,quantile(wines$total.sulfur.dioxide,0.99))) 

p8 = ggplot(data=wines,aes(x=density))+
  geom_histogram(binwidth=.0005)+
  xlim(c(0.9871,quantile(wines$density,0.99))) 

p9 = ggplot(data=wines,aes(x=pH))+
  geom_histogram(binwidth=.02)

p10 = ggplot(data=wines,aes(x=sulphates))+
  geom_histogram(binwidth=.02)

p11 = ggplot(data=wines,aes(x=alcohol))+
  geom_histogram(binwidth=.2)

p12 = ggplot(data=wines,aes(x=quality))+
  geom_bar()

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol=3)
```

This is now a bit clearer. I notice citric acid is normally distributed but has 
a spike at around 0.48. This is likely due to standard from the wine industry.
We see that many of the distributions are normallydistributed while some are 
not. I read the notes to gain a better understanding of each attribute. We will 
now target specific attributes to explore further.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_quality}
ggplot(data=wines,aes(x=quality))+
  geom_histogram(binwidth=1)

table(wines$quality)
```

I note that the lowest quality rating is a 3 (20 wines) and the highest quality 
rating is a 9 (5 wines). This variable can probably be turned into an ordinal 
factor variable for analysis. We will do this later.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_alcohol}
wines$alcohol.bucket<-cut(wines$alcohol,c(8,9,10,11,12,13,14))

ggplot(data=subset(wines,!is.na(alcohol.bucket)),aes(x=alcohol.bucket))+
  geom_bar()

summary(wines$alcohol)
```

To make the data more meaningful I decide to create a new variable called
alcohol.bucket which will group wines according to alcohol content. I decide to
create 1% buckets so 8-9%, 9-10% etc. Our resulting plot shows that most wines 
contain 9-10% of alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_residual_sugar_outliers}
grid.arrange(ggplot(wines, aes(x=1, y=residual.sugar ) )+ 
               geom_jitter(alpha=0.1 ) +
               geom_boxplot(alpha=0.2, color = 'blue' ),
             ggplot(wines,aes(x=residual.sugar))+ 
                   geom_histogram(bins=30),ncol=2)

summary(wines$residual.sugar)
```

Residual sugar contains a number of outliers above 20. Outliers are values much
greater than the norm. As we can see from the boxplot the mode is only about 5
but there are a few values over 60. As outliers can skew our results we will
remove them in our next plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_residual_sugar}
p1 = ggplot(data=wines,aes(x=residual.sugar))+
  geom_histogram(binwidth=0.2)+
  xlim(c(0.6,30))

p2 = ggplot(data=wines,aes(x=residual.sugar))+
  geom_histogram(binwidth=0.1)+
  scale_x_log10(limits=c(0.6,30))

grid.arrange(p1, p2, ncol=1)

sum(wines$residual.sugar>45)
```

Residual sugar is the amount of sugar in the wine after fermentation has
stopped. It appears that most white wines are not particuarly sweet. As
indicated in the notes, wine is considered sweet only if it contains over 45
grams/liter of residual sugar. In this case only 1 wine one contains more than 
45 grams/liter.

The original plot shows a sizeable number of wines greater than the mean. To 
gain a better understanding we replot this using a log10 transformation. As
previously stated we also remove outliers by limiting residual sugar to a 
maximum of 30.

The resulting plot shows a bimodial distribution. It is interesting to find that
there are as much if not more wines with residual sugar greater than 
3 grams/liter than there are less than 3 grams/liter.

# Univariate Analysis

### What is the structure of your dataset?

The dataset contains 4898 rows and 13 variables. One variable is just a running
number which we will ignore. Another variable is the quality rating given by 
wine tasters and that is our output variable. The other 11 are chemical 
properties of the wine.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in the dataset is the quality rating given to each
wine. This is the outcome variable and we are trying to determine which
attributes are best for predicting wine quality.

### What other features in the dataset do you think will help support your\
investigation into your feature(s) of interest?

At this stage most of the 11 attributes can potentially influence the quality 
rating of the wine. Googling wine tasting, I find that an often repeated 
important criteria is balance, that is, the balance of  various components in 
the wine such as fruity, sweet and sour, bitter and earthy characteristics. 
Each of these attributes can potentially affect this balance. 

Having said that, my exploration so far gives me a better understanding of each
attribute. I think alcohol and residual sugar may help support my investigation. 

### Did you create any new variables from existing variables in the dataset?

Yes I created an alcohol.bucket variable where each bucket is 1% of alcohol. I 
find grouping wines into these buckets makes it easier to spot trends in the 
plots.

### Of the features you investigated, were there any unusual distributions? Did\
you perform any operations on the data to tidy, adjust, or change the form of\
the data? If so, why did you do this?

I noticed residual sugar was long-tailed with a substantial amount of data in
the long-tail section even after removing the top 1%. To make it easier to 
observe and understand the data I decided to apply a log 10 tranformation. By
doing this we were able to observe a bimodial distribution which we otherwise
wouldn't have been able to observe.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_corr}
wines = read.csv('wineQualityWhites.csv')
w <- wines[,c(2:13)]
cor(w,method="pearson")
```

I start by exploring the correlations between variables by outputting a 
correlation matrix of all 12 variables. No attributes are highly correlated with
quality. The two highest correlated attributes with a medium correlation 
are alcohol (0.43) and density (-0.3).

Of the other attributes the following have only slight (low) correlation with 
quality: fixed acidity (-0.11), volatile acidity (-0.19), chlorides (-0.2)
and total sulfur dioxide (-0.17).

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_corr_plot}
ggpairs(w, columns=1:12,
        columnLabels=c("f.acd","v.acd","c.aci","r.sug",
                                      "chlor","fsd","tsd","dens",
                                      "pH","sulph","alcl","qual"),
        upper = list(continuous = wrap("cor", size = 3)))+
  theme(axis.text = element_blank())
```

I perform a correlation plot of all our variables to get a quick visual overview
of the relationships.

The following attributes have the following correlation with alcohol:

* fixed acidity (very low)
* volatile acidity (very low)
* chlorides (low)
* total sulfur dioxide (medium)
* density (high)

The following attributes have the following correlation with density:

* fixed acidity (low)
* volatile acidity (very low)
* chlorides (low)
* total sulfur dioxide (medium)
* alcohol (high)

This information is useful because as we build our model, we have a better idea
of which  variables to include or not include in our model. Variables that are 
correlated with each other may not add much to the model. On the other hand 
variables that have some correlation with quality but not with each other may be
a good candidate.

We will now look closer at correlations betwen specific variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_alcohol}
w$quality.cat <- factor(w$quality)

ggplot(data=w,aes(x=quality.cat,y=alcohol))+
  geom_jitter(alpha=1/12)+
  geom_boxplot(alpha=1/5,color='blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

cor(w$alcohol, w$quality)
```

I decide to change the quality attribute into an ordinal factor (category) so 
that boxplots / scatterplots of each quality category can be plotted next to 
each other and we can easily compare the amount of alcohol for each one. The 
scatterplot will give you an idea of the number of wines under each quality 
category which the boxplot does not provide. The mean is also calculated as
denoted by a red star.

In this plot I find a distinct trend from quality level 5 onwards that as 
alcohol increases then so does the quality rating.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_density}
ggplot(data=w,aes(x=quality.cat,y=density))+
  geom_boxplot(alpha=1/5,color='blue')+
  geom_jitter(alpha=1/12)+
  stat_summary(fun.y = "mean", 
             geom = "point", 
             color = "red", 
             shape = 8, 
             size = 4)+
  ylim(c(0.9871,quantile(w$density,0.99))) 

cor(w$density, w$quality)
```

I do the same for density vs quality. As in the univariate analysis section I 
remove the top 1% to remove outliers. Here we see a trend that when density 
decreases, quality increases.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_fixed_acidity}
ggplot(data=w,aes(x=quality.cat,y=fixed.acidity))+
  geom_boxplot(alpha=1/5,color='blue')+
  geom_jitter(alpha=1/12)+
  stat_summary(fun.y = "mean", 
             geom = "point", 
             color = "red", 
             shape = 8, 
             size = 4)+
  ylim(c(3.8,quantile(w$fixed.acidity,0.99)))

cor(w$fixed.acidity, w$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_volatile_acidity}
ggplot(data=w,aes(x=quality.cat,y=volatile.acidity))+
  geom_boxplot(alpha=1/5,color='blue')+
  geom_jitter(alpha=1/12)+
  stat_summary(fun.y = "mean", 
             geom = "point", 
             color = "red", 
             shape = 8, 
             size = 4)+
  ylim(c(0.08,quantile(w$volatile.acidity,0.99))) 

cor(w$volatile.acidity, w$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_chlorides}
ggplot(data=w,aes(x=quality.cat,y=chlorides))+
  geom_boxplot(alpha=1/5,color='blue')+
  geom_jitter(alpha=1/12)+
  stat_summary(fun.y = "mean", 
             geom = "point", 
             color = "red", 
             shape = 8, 
             size = 4)+
  ylim(c(0,quantile(w$chlorides,0.99)))

cor(w$chlorides, w$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_total_sulfur_dioxide}
ggplot(data=w,aes(x=quality.cat,y=total.sulfur.dioxide))+
  geom_boxplot(alpha=1/5,color='blue')+
  geom_jitter(alpha=1/12)+
  stat_summary(fun.y = "mean", 
           geom = "point", 
           color = "red", 
           shape = 8, 
           size = 4)+
  ylim(c(0,quantile(wines$total.sulfur.dioxide,0.99)))

cor(w$total.sulfur.dioxide, w$quality)
```

The above 4 plots of fixed acidity, volatile acidity, chlorides and total sulfur
dioxide all have low correlation to quality but I have included them to see if
I could find any interesting trends. Here are my observations:

Fixed acidity - The plot is almost flat with only a slight decreasing trend
that when fixed acidity decreases, quality increases.

Volatile acidity - Overall the trend is somewhat like a sine curve. There is a
increase in quality when there is a drop in volatile acididity between the two
most populous buckets 5 and 6 but then trend goes opposite direction.

Chlorides - It is interesting to see that the amount of chlorides do seem to
suggest a difference in rating. In the quality buckets 5, 6 and 7 where there
are the most number of wines, we can see a noticeable trend that as chlorides 
decreased then quality did increase.

Total Sulfur dioxide - There is a distinct trend between the 3 most populous
alcohol buckets 5, 6 and 7 that when total sulfur dioxide decreases, quality
rating increases. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_alcohol_sugar_density}
w$alcohol.bucket<-cut(w$alcohol,c(8,9,10,11,12,13,14))

ggplot(data=subset(w,!is.na(alcohol.bucket)),
            aes(x=alcohol.bucket,y=density))+
  geom_boxplot(alpha=1/5,color='blue')+
  geom_jitter(alpha=1/12)+
  stat_summary(fun.y = "mean", 
           geom = "point", 
           color = "red", 
           shape = 8, 
           size = 4)+
  ylim(c(0.9871,quantile(w$density,0.99))) 

cor(w$alcohol, w$density)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_sugar_density}
ggplot(data=w,aes(x=residual.sugar,y=density))+
  geom_jitter(alpha=1/10)+
  xlim(c(0.6,quantile(w$residual.sugar,0.99))) +
  ylim(c(0.9871,quantile(w$density,0.99))) 

cor(w$residual.sugar, w$density)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_tsd_density}
ggplot(data=w,aes(x=total.sulfur.dioxide,y=density))+
  geom_jitter(alpha=1/10)+
  xlim(c(0,quantile(wines$total.sulfur.dioxide,0.99)))+
  ylim(c(0.9871,quantile(w$density,0.99))) 

cor(w$total.sulfur.dioxide, w$density)
```

Alcohol and residual sugar have a high correlation to density so I take a closer 
look at their relationships. Looking at the plots I can see a distinct trend 
that density decreases when alcohol increase and residual sugar decrease and
vice versa. This is probably due to the heaviness / lightness of each component
which will affect the density.

I also take a look at total sulfur dioxide against density as they have a
moderate correlation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_sugar_alcohol}
ggplot(data=w,aes(x=residual.sugar,y=alcohol))+
  geom_jitter(alpha=1/10)+
  xlim(c(0.6,quantile(w$residual.sugar,0.99))) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_tsd_alcohol}
ggplot(data=w,aes(x=total.sulfur.dioxide,y=alcohol))+
  geom_jitter(alpha=1/10)+
  xlim(c(0,quantile(w$total.sulfur.dioxide,0.99))) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_cholorides_alcohol}
ggplot(data=w,aes(x=chlorides,y=alcohol))+
  geom_jitter(alpha=1/10)+
  xlim(c(0,quantile(w$chlorides,0.99))) 
```

Residual sugar, total sulfur dioxide and chlorides are all moderately correlated 
with alcohol so I decide to take a closer look these variables.

With residual sugar, looking at the right side of the plot, we can see that the
more residual sugar the lower the alcohol. However the opposite does not appear 
to be true. The less residual sugar does not necessarily mean greater amounts
of alcohol as we can see a good spread of data in the lower residual sugar 
areas.

Total sulfur dioxide also has an interesting pattern. Where total sulfur dioxide
is around the middle, alcohol is quite evenly spread out as can be seen by the
amount of dots at between 100-150 of total sulfur dioxide. However greater than
150 alcohol starts to drop.

With chlorides between around 0.01 to 0.025 alcohol content tends to be higher. 
and between around 0.05 to 0.065 alcohol content tends to be lower. In the
middle at around 0.03 alcohol tends to be evenly spread out. Also it is
interesting that from 0.065 and greater alcohol remains low at between 8-10%

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The attributes with the highest correlation to quality is alcohol followed by
density with a low to medium correlation. No attributes were highly correlated
to quality.

There were several other attribute that had quite low correlations with quality. 
These were total fixed acidity, volatile acidity, chlorides and total
sulfur dioxide.

I know from reading up on wine tasting that there are many components at play 
when rating wine - balance being one of them, that is the various components 
within the wine that make up the taste. With this in mind I didn't want to 
discount any of the attributes unnecessarily. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I observed that residual sugar and alcohol are highly correlated with density
probably due to the characteristics of each component. When there is more sugar
in the wine (heavier) density increased and when there is more alcohol in the
wine (lighter) density decreased.

I observed some not straightforward trends with alcohol vs residual sugar. 
When residual sugar was high it meant lower amounts of alcohol. However when
residual sugar was low it did not necessarily mean alocohol was high. Likewise
with alcohol vs total sulfur dioxide, when total sulfur dioxide was higher, 
alcohol tended to be lower. But around the mean and less, alcohol was more
evenly spread out. These non linear trends could be due to chemical thresholds.

Alcohol vs chlorides was also interesting in that the majority of wines had 
around 0.05 of chlorides. Alcohol tended to be evenly spread but then when 
greater than the mean alcohol starts to drop. Again it could be some chemical
thresholds at play.

I also found that on some plots such as volatile acidity vs quality it showed
trends in the quality buckets 5, 6 and 7 which decrease in volatile acidity.
These buckets also happened to be the most populous buckets. So even though
the extreme buckets on either end bucks the trend, it may not matter as much 
as these buckets have a much smaller number of wines.

### What was the strongest relationship you found?

The strongest relationship to our feature variable quality is alcohol followed 
by density. However alcohol and density have a high correlation between 
themeselves. Residual sugar is also hghly correlated with density.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_density_sugar_alcohol}
ggplot(data=subset(w,!is.na(alcohol.bucket)),
       aes(x=residual.sugar,y=density,color=alcohol.bucket))+
  geom_jitter(alpha=1/2)+
  geom_smooth(method = "lm", se = FALSE,size=.8)+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Alcohol',
    override.aes = list(alpha = 1, size = 2)))+
  xlim(c(0.6,quantile(wines$residual.sugar,0.99)))+
  ylim(c(0.9871,quantile(w$density,0.99)))
```

I know from previous analysis that two highest correlated attributes to quality
is density and alcohol so that is my starting point. 

I first look at the relationship between density, alcohol and residual sugar
and this plot proves what we already know, that all three variables are highly 
collerated. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_quality_buckets}
table(w$quality)
```

I'm interested to tie this back to our main feature variable quality. I decide 
to regroup the quality buckets by combining two lowest ratings (3 and 4) into 
one bucket and the two highest ratings (8 and 9) into another bucket. The reason
I do this is because previously I found that, for some variables, trends existed
only in the 5, 6, 7 quality rating buckets which was also the most populous. The
buckets at the end in comparison had a lot fewer wines. I wondered if I combined
these end buckets, whether it would more clearly show any trends. (For example
the plots chlorides vs quality and total sulfur dioxide vs quality exhibited 
this trend.)

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_density_sugar_alcohol_facets}
w$quality.cat2 <-cut(w$quality,c(3,5,6,7,9))

ggplot(data=subset(subset(w,!is.na(alcohol.bucket)),!is.na(quality.cat2)),
       aes(x=residual.sugar,y=density,color=alcohol.bucket))+
  geom_jitter(alpha=1/2)+
  geom_smooth(method = "lm", se = FALSE,size=.6)+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Alcohol',
    override.aes = list(alpha = 1, size = 2)))+
  xlim(c(0.6,quantile(wines$residual.sugar,0.99)))+
  ylim(c(0.9871,quantile(w$density,0.99)))+
  facet_wrap(~quality.cat2)
```

I replot the previous graph using facet_wrap on the new quality buckets. In this
case we can clearly see the interplay between the 4 variables. For example 
comparing the 7,9 bucket with the 3,5 bucket we can see that generally lower 
density, higher alcohol meant a higher rating. Residual sugar also seems to be 
less in the 7,9 bucket. Also as we found previously less residual sugar did not 
particularly mean a higher rating and that fact was also reflected in this plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_tsd_alcohol_density}
w$total.sulfur.dioxide.cat <-cut(w$total.sulfur.dioxide,c(9,108,134,167,440))

p1 = ggplot(data=subset(subset(w,!is.na(quality.cat2)),
                      !is.na(total.sulfur.dioxide.cat)),
            aes(x=quality.cat2,y=density,color=total.sulfur.dioxide.cat))+
  geom_jitter(alpha=1/2)+
  ylim(c(0.9871,quantile(w$density,0.99)))+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Total Sulfur Dioxide',
    override.aes = list(alpha = 1, size = 2)))

p2 = ggplot(data=subset(subset(w,!is.na(quality.cat2)),
                      !is.na(total.sulfur.dioxide.cat)),
            aes(x=quality.cat2,y=alcohol,color=total.sulfur.dioxide.cat))+
  geom_jitter(alpha=1/2)+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Total Sulfur Dioxide',
    override.aes = list(alpha = 1, size = 2)))

grid.arrange(p1, p2, ncol=1)
```

Total sulfur dioxide is moderately correlated with both alcohol and density and
lowly correlated to quality. I want to take a look at these three variables 
together. I cut total sulfur dioxide into buckets based on its interquartile 
range. I also decide to plot total sulfur dioxide vs alcohol and total sulfur 
dioxide vs density on top of each other so I can compare them.

The two plots both exhbit the expected trends. Visually the alcohol plot looks 
a bit cleaner probably because alcohol is more correlated to quality than 
density.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_chlorides_alcohol_density}
w$chlorides.cat <-cut(w$chlorides,c(.009,.036,.043,.05,.346))

p1 = ggplot(data=subset(subset(w,!is.na(quality.cat2)),!is.na(chlorides.cat))
            ,aes(x=quality.cat2,y=density,color=chlorides.cat))+
  geom_jitter(alpha=1/2)+
  ylim(c(0.9871,quantile(w$density,0.99)))+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Chlroides',
    override.aes = list(alpha = 1, size = 2)))

p2 = ggplot(data=subset(subset(w,!is.na(quality.cat2)),!is.na(chlorides.cat)),
            aes(x=quality.cat2,y=alcohol,color=chlorides.cat))+
  geom_jitter(alpha=1/2)+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Chlorides',
    override.aes = list(alpha = 1, size = 2)))

grid.arrange(p1, p2, ncol=1)
```

I do the same with chlorides as it is somewhat correlated with density, alcohol
and quality. I cut chlorides into buckets based on its interquartile range.
We can see quite clear trends especially the plot with alcohol. This plot 
doesn't really tell me anything new that I don't already know.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_va_alcohol}
w$volatile.acidity.cat <-cut(w$volatile.acidity,c(.08,.21,.26,.32,1.1))

ggplot(data=subset(subset(w,!is.na(quality.cat)),!is.na(volatile.acidity.cat)),
       aes(x=quality.cat, y=alcohol,color=volatile.acidity.cat))+
  geom_jitter(alpha=1/2)+
  scale_color_brewer(type = 'seq',
    guide = guide_legend(title = 'Volatile Acidity',
    override.aes = list(alpha = 1, size = 2)))
```

Previously I found there was a clear trend in volatile acidity vs quality in the
middle ratings. I now want to see if volatile acidity strengthens the alcohol vs
quality relationship. The plot appears to show that it does. Volatile acidity is
more evident in the lower alcohol and lower quality ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_sugar_alcohol2}
w$residual.sugar.cat <-cut(w$residual.sugar,c(0,4,40))

ggplot(data=subset(subset(w,!is.na(quality.cat)),!is.na(residual.sugar.cat)),
       aes(x=quality.cat,y=alcohol,color=residual.sugar.cat))+
  geom_jitter(alpha=1/2)+
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Residual Sugar',
    override.aes = list(alpha = 1, size = 2)))
```

In an earlier plot I found that residual sugar when transformed using a log10
scale displayed a bimodial plot which suggests it would lend itself well to 2 
categories of less sugar and more sugar. I decide to explore this further by 
cutting residual sugar into 2 buckets along these lines and comparing it with 
alcohol and quality.

In this plot I can see trends between the three variables as they are all 
somewhat correlated. However it is hard to tell how much residual sugar actually
strengthens the relationship if any.

It is surprising to find that quality buckets 3 and 4 and buckets 8 and 9
have residual sugar from both buckets in about even amounts. You would have 
expected to see less residual sugar equate to higher ratings. And this is 
also despite alcohol being higher in the buckets 8 and 9 compared to 3 and 4.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_linear_model}
m1 <- lm(quality ~ alcohol , data=w)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + residual.sugar)
m4 <- update(m3, ~ . + density)
m5 <- update(m4, ~ . + chlorides)
mtable(m1,m2,m3,m4,m5,sdigits=3)

```

This linear model accounts for 26.4% of the variance in the quality rating of 
white wines. I note that volatile acidity which only had some correlation
with quality but was lowly correlated with alcohol did increase the R-squared 
value in the model.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Volatile acidity does seem to strengthen the alcohol vs quality relationship.
We can quite clearly see a trend that as you move through the quality rating 
buckets from low to high that there are more and more alcohol and less and less
volatile acidity.

Variables which were correlated with each other, when plotted, displayed the
expected trends but was sometimes hard to determine if they actually 
strengthened each other. Perhaps they did a little but it is hard to work that 
out visually.

### Were there any interesting or surprising interactions between features?

The low quality rating bucket (3 and 4) and high quality rating bucket (8 and 9)
almost had the same spread of residual sugar. This bucked the trend of the 
middle buckets where there the less residual sugar meant more alcohol and higher
quality ratings. This might suggest that there are other elements at play here
where it is not just a straight residual sugar / alcohol / quality relationship.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes I created a linear model from my dataset. I used alcohol, the most 
correlated variable with quality as my starting point. I wasn't sure if density, 
the second most correlated variable with quality would make alot of difference 
to the R-squared. In the end it made a slight difference. Volatile acidity a 
lowly correlated variable made a difference as it was not correlated with 
alcohol. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_One}
w$quality.cat <- factor(w$quality)

ggplot(data=w,aes(x=quality.cat,y=alcohol))+
  geom_jitter(alpha=1/12)+
  geom_boxplot(alpha=1/5,color='blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  scale_y_continuous(name='Alcohol %')+
  scale_x_discrete(name='Quality (1=worst, 10=best)')+
  ggtitle("Relationship between Alcohol and Quality")
```

### Description One

One of my first plots is to explore alcohol against quality. This plot is 
interesting because there are clear trends between quality buckets 5, 6 and 7. 
As alcohol increases the quality rating also increases. This can be seen in the 
following buckets.

* Bucket 5 - most wines have low levels of alcohol and there are not that many 
with high levels. 
* Bucket 6 - is more evenly spread out
* Bucket 7 - is the opposite with more wines with higher levels of alcohol. 

What I am also interested to explore in the future are the wines that condradict
this trend. There are some wines in quality bucket 8 that have about 9% alcohol.
How did these wines with low alcohol receive such high ratings. Likewise there 
is a smattering of medium alcohol wines that received low quality ratings of 3 
and 4. Why did these wines rate so badly?

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Two}
w$alcohol.bucket<-cut(w$alcohol,c(8,9,10,11,12,13,14),
                      labels=c('8','9','10','11','12','13'))
w$quality.cat2 <-cut(w$quality,c(3,5,6,7,9),
                     labels=c("Low (3-4)","Medium Low (5)","Medium High (6)","High (7-9)"))

ggplot(aes(x=alcohol.bucket,y=volatile.acidity),
       data=subset(subset(w,!is.na(quality.cat2)),!is.na(alcohol.bucket)))+
  geom_boxplot(aes(fill=quality.cat2))+
  scale_fill_brewer(type='seq',
                    guide=guide_legend(title='Quality'))+
  scale_y_continuous(name='Volatile Acidity (g/dm^3)',limits=c(0.08,quantile(w$volatile.acidity,0.90)))+
  scale_x_discrete(name='Alcohol (%)')+
  ggtitle('Alcohol and Volatile Acidity Impact on Quality')
```

### Description Two

This plot was redesigned from earlier using boxplots instead of a scatterplot.

This plot presents some interesting and not so straight-forward trends. For 
example it shows that in alcohol buckets 8,9 and 10 the lower the volatile
acidity the higher the quality rating. However in the higher alochol buckets 
11,12 and 13 volatile acidity seems to be quite even amongst the different 
quality ratings.

Alcohol on the other hand, as it increases, then generally so does quality as 
can be seen by the darker bars in the higher alcohol buckets. The relationship 
between volatile acidity and alcohol is that as alcohol increases then generally
so does volatile acidity.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Three}
w$alcohol.bucket<-cut(w$alcohol,c(8,9,10,11,12,13,14),
                      labels=c('8%','9%','10%','11%','12%','13%'))
w$quality.cat2 <-cut(w$quality,c(3,5,6,7,9))

quality_names <- list(
  'Q1'="Quality (3-4 Low)",
  'Q2'="Quality (5 Medium -)",
  'Q3'="Quality (6 Medium +)",
  'Q4'="Quality (7-9 High)"
)

quality_labeller <- function(variable,value){
  return(quality_names[value])
}

ggplot(data=subset(subset(w,!is.na(alcohol.bucket)),!is.na(quality.cat2)),
       aes(x=residual.sugar,y=density,color=alcohol.bucket))+
  geom_jitter(alpha=1/2)+
  geom_smooth(method = "lm", se = FALSE,size=.6)+
  scale_color_brewer(type = 'div', palette='RdYlBu', direction=-1,
    guide = guide_legend(title = 'Alcohol'))+
  scale_y_continuous(name='Density (g/cm^3)',
                     limits=c(0.9871,quantile(w$density,0.99)))+
  scale_x_continuous(name='Residual Sugar (g/dm^3)',
                     limits=c(0.6,quantile(wines$residual.sugar,0.99)))+
  facet_wrap(~quality.cat2, labeller=quality_labeller)+
  ggtitle('Alcohol, Density and Residual Sugar impact on Quality')
```

### Description Three

This plot provides a good overview of the interplay between 3 attributes of wine
- alcohol, density and residual sugar to quality. There is visible correlation 
between alcohol, density and quality rating - the higher the rating the higher 
the alcohol and the lower the density. When adding residual sugar to the mix we 
can see that when wines are more highly rated they tend to have less sugar. 
However when there is less sugar it didn't necessarily mean higher ratings as 
all 4 quality buckets had a good porportion of wines with little residual sugar.

------

# Reflection

Which direction to take in my exploration was the hardest part of this project. 
I wasn't very familiar with the chemical properties of wine to begin with so I 
had to perform some research on wine tasting using Google. In hindsight having 
a good understanding of the domain in which you are investigating would greatly 
help guide which direction to take when exploring the data as it is so open 
ended.

I found some interesting trends when performing some of the plots. Sometimes 
this lead to further questions which you will try and follow through with 
further plots. This was quite rewarding especially when the line of enquiry 
finally lead to insights that wasn't at first apparent. In some lines of enquiry
there were results that followed the expected trends but there were also some 
parts of the data that bucked the trend. For exampleless residual sugar 
generally meant more alcohol and a higher quality rating but in the lowest and
highest quality ratings buckets  the amount of sugar was quite evenly spread 
out. This is probably due to certain thresholds that when sugar is greater or 
less than certain values they no longer affect the quality rating.

Generally I found that the exploratary data analysis process is not a linear 
one. Many times I would revisit univariate or bivariate analysis when performing
the multivariate section. One discovery led to further questions which 
subsequently meant revisiting existing plots or plotting new ones.

For the future I think there is scope to delve deeper into some of the analyses.
For example where there are attributes with certain parts that bucked the trend
it would be interesting to isolate these areas and see how other variables 
interact with it. Maybe trends can be established within specific subsets of 
data.



